---
title: "MLB Player Analysis"
author: "Collin Martin, Lukas Juranek"
date: "UC3M, 2024"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

```{r}
rm(list=ls())
set.seed(123)
# Data Manipulation
library(tidyverse)  # Comprehensive data manipulation and visualization
library(dplyr)      # Data manipulation (part of tidyverse)
library(stringr)    # String manipulation
library(reshape2)   # Data reshaping and melting
# Visualization
library(ggplot2)    # Data visualization
library(corrplot)   # Correlation matrix visualization
library(ggrepel)    # Improved text label placement
library(hrbrthemes) # Ggplot themes
library(viridis)    # Color scales for ggplot2
# Statistical Analysis and Clustering
library(factoextra) # PCA and clustering visualization
library(cluster)    # Clustering algorithms
library(mclust)     # Model-based clustering
library(nFactors)   # Determining number of factors
library(insight)    # Extract and interpret model information
library(caret)      # Machine learning

pitching_df <- read.csv("MLB_pitching_stats_ALL_PITCHING.csv")
head(pitching_df)
dim(pitching_df)
```

# pitching_df Preprocessing

## Visualization

```{r}

# Count the number of 0 values in each column
colSums(pitching_df == 0)
par(mar = c(5, 10, 4, 2))  # Increase left margin to fit labels
barplot(colSums(pitching_df == 0)[colSums(pitching_df == 0) >= 2], las=2, horiz = T)


# Plot games played
ggplot(pitching_df, aes(x = gamesPlayed)) +
  geom_histogram(binwidth = 1, fill = "darkblue", color = "black")+
  geom_vline(xintercept = 5, color = "red", linetype = "dashed")

```

```{r}

# Correct pitching_df types for some character columns
pitching_df <- pitching_df %>%
  mutate(across(c("winPercentage", "runsScoredPer9", "babip", "strikeoutsPer9", "baseOnBallsPer9", "homeRunsPer9", "hitsPer9", "strikesoutsToWalks", "pitchesPerInning", "walksPerStrikeout", "stolenBasePercentage", "era", "whip", "groundOutsToAirouts", "winPercentage", "winPercentage", "strikeoutWalkRatio", "strikeoutsPer9Inn", "walksPer9Inn", "hitsPer9Inn"), as.numeric))

pitching_df <- pitching_df %>% filter(year == 2024)
pitching_df <- pitching_df %>% filter(gamesPlayed > 5)

pitching_df$gidpOpp <- ifelse(pitching_df$battersFaced > 0, pitching_df$gidpOpp / pitching_df$battersFaced, NA)

pitching_df$swingAndMisses <- ifelse(pitching_df$battersFaced > 0, pitching_df$swingAndMisses / pitching_df$battersFaced, NA)

pitching_df$ballsInPlay <- ifelse(pitching_df$battersFaced > 0, pitching_df$ballsInPlay / pitching_df$battersFaced, NA)

# Define the columns to mark for deletion
columns_to_delete <- c("X",
  "playerId", "year", "playerName", "type", "rank", "playerFirstName",
  "playerLastName", "playerUseName", "playerInitLastName", "teamId",
  "runSupport", "teamAbbrev", "teamName", "leagueName", "leagueId",
  "position", "primaryPositionAbbrev", "ops",
  "strikesoutsToWalks", "stolenBases", "caughtStealing", "doubles",
  "triples", "gidp", "wildPitches", "balks", "pickoffs", "totalSwings",
  "pitchesPerInning", "walksPerStrikeout","lineOuts", "airOuts",
  "catchersInterference", "sacBunts", "sacFlies",
  "groundOuts", "strikeoutWalkRatio", "popOuts", 'lineOuts', 'groundOuts',
  "winningPercentage",
  "wins","losses", "atBats", "numberOfPitches",
  "gamesPitched", "intentionalWalks",
  "strikeoutsPer9Inn", "baseOnBallsPer9",
  "hitsPer9Inn","gidp",
  "hitBatsmen", "groundIntoDoublePlay", "gamesPlayed","stolenBasePercentage"
)
# Added : groundintodoubleplay, gamesplayed
# pitching_df Kept: "popHits", "lineHits","groundHits", "flyOuts","wins", "losses"

# Remove selected columns
pitching_df <- pitching_df %>%
  select(-one_of(columns_to_delete))
# Only keep numerical pitching_df
pitching_df.num <- pitching_df %>%
  select_if(is.numeric)

corr_matrix <- cor(pitching_df.num, use = "complete.obs")

corrplot(corr_matrix, 
         method = "color",
         col = colorRampPalette(c("blue", "white", "red"))(200), 
         type = "upper",
         tl.col = "black",
         tl.srt = 45,
         addCoef.col = "black",
         number.cex = 0.35,
         tl.cex = 0.5, 
         
         number.digits = 3 # Set precision for display
)
```

Find and print pairs with correlation 1

```{r}
# Find and print pairs with correlation 1
correlated_columns <- which(corr_matrix == 1, arr.ind = TRUE)
print("Pairs with correlation 1:")
for (i in 1:nrow(correlated_columns)) {
  row <- correlated_columns[i, 1]
  col <- correlated_columns[i, 2]
  
  # Only print pairs where the row index is less than the column index to avoid duplicates
  if (row < col) {
    print(paste(colnames(cor_matrix)[row], colnames(cor_matrix)[col]))
  }
}

```

# Handling missing values and outliers

```{r}
# Replace NA values with the mean of the column
pitching_df.num <- pitching_df.num %>%
  mutate(across(everything(), ~ifelse(is.na(.), mean(., na.rm = TRUE), .)))

```

Show the distributions of all variables

```{r}
data_long <- pitching_df.num %>% 
  pivot_longer(everything(), names_to = "variable", values_to = "value")

ggplot(data_long, aes(x = value)) +
  geom_density(fill = "blue", alpha = 0.4) +
  facet_wrap(~ variable, scales = "free") +
  labs(title = "Distributions of All Variables", x="", y="") +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    strip.text = element_text(size = 7)
)
ggsave("distributions_all.png", plot = last_plot(), width = 8, height = 6, dpi = 300)

# Box plot of all variables
scaled_data <- scale(sqrt(pitching_df.num))
boxplot(scaled_data, las=2)
```

ERA histrogram

```{r}
p <- ggplot(pitching_df.num, aes(x = pitching_df$era)) +
      geom_histogram(bins = 50, fill = "darkblue", color = "black", alpha = 0.7) +
      # geom_boxplot(fill = "darkblue", color = "black", alpha = 0.7) +
      labs(title = paste("Histogram of era"), x = "Era", y = "Frequency") +
      theme_minimal()
    # Print the plot to display it
print(p)
ggsave("histogram_era.png", plot = last_plot(), width = 8, height = 6, dpi = 300)
```

# pitching_df Analysis

### Correlation of important variables

```{r}
cor_matrix <- cor(pitching_df[, c("babip", "obp", "slg", "strikeoutsPer9", "homeRunsPer9", "hitsPer9", "era", "whip")])
corrplot(cor_matrix, method = "color", addCoef.col = "black", number.cex=0.7,col = colorRampPalette(c("blue", "white", "red"))(200))

# Correlations for starting pitchers
starter_pitching_df <- pitching_df[pitching_df$gamesStarted > 4, ]
cor_matrix_starters <- cor(starter_pitching_df[, c("era", "whip", "strikeoutsPer9", "homeRunsPer9", "hitsPer9", "baseOnBalls", "inningsPitched", "gamesStarted")])
corrplot(cor_matrix_starters, method = "color", tl.cex = 0.8, addCoef.col = "black", number.cex=0.7, col = colorRampPalette(c("blue", "white", "red"))(200))

# Correlations for relief pitchers
reliever_pitching_df <- pitching_df[pitching_df$gamesFinished > 4, ]
cor_matrix_relievers <- cor(reliever_pitching_df[, c("era", "whip", "strikeoutsPer9", "homeRunsPer9", "hitsPer9", "baseOnBalls", "gamesFinished", "saves", "saveOpportunities", "holds", "blownSaves")])
corrplot(cor_matrix_relievers, method = "color", tl.cex = 0.8, addCoef.col = "black", number.cex=0.7, col = colorRampPalette(c("blue", "white", "red"))(200))

```

## Plots of important variables

```{r}
ggplot_era <- ggplot(pitching_df, aes(x = era, y = whip)) +
  geom_point(aes(color = gamesStarted)) +
  labs(title = "ERA vs WHIP", x = "ERA", y = "WHIP") +
  theme_minimal() +
  scale_color_gradient(low = "blue", high = "red")
ggplot_era # Lower whip and era values correlate to lower values with the direct other variable
ggsave("era_whip.png", plot = last_plot(), width = 8, height = 6, dpi = 300)

ggplot_strikouts <- ggplot(pitching_df, aes(x = strikeoutsPer9, y = walksPer9Inn)) +
  geom_point(aes(color = battersFaced)) +
  labs(title = "Strikeouts per 9 vs Home Runs per 9", x = "Strikeouts per 9", y = "Home Runs per 9") +
  theme_minimal() +
  scale_color_gradient(low = "blue", high = "red")
ggplot_strikouts# Values seem uncorrelated
ggsave("strikeouts_walks.png", plot = last_plot(), width = 8, height = 6, dpi = 300)
# Plot data for Starting pitchers
ggplot(starter_pitching_df, aes(x = battersFaced, y = era)) +
  geom_point(aes(color = battersFaced)) +
  labs(title = "Strikeouts Over Time", x = "Batters Faced", y = "Strike Outs") +
  theme_minimal() +
  scale_color_gradient(low = "blue", high = "red")

# plot data for Relievers
ggplot(reliever_pitching_df, aes(x = battersFaced, y = era)) +
  geom_point(aes(color = era)) +
  labs(title = "Strikeouts Over Time", x = "Batters Faced", y = "Strike Outs") +
  theme_minimal() +
  scale_color_gradient(low = "blue", high = "red")

# Assuming you have two previously created data frames : starter_pitching_df and reliever_pitching_df

# Add a new column to label the data as "Starter" or "Reliever"
starter_pitching_df$Position <- "Starter"
reliever_pitching_df$Position <- "Reliever"

# Combine the two data frames into one
combined_pitching_df <- rbind(starter_pitching_df, reliever_pitching_df)

# Now, plot both data sets with two boxplots side by side
box_plot_era <-ggplot(combined_pitching_df, aes(x = Position, y = era, fill = Position)) +
  geom_boxplot() +
  scale_fill_viridis(discrete = TRUE, alpha = 0.6) +
  geom_jitter(color = "black", size = 0.4, alpha = 0.9) +
  theme_ipsum() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 11),
    axis.text.x = element_text(angle = 0, hjust = 0.5)  # Adjust x-axis labels for readability
  ) +
  coord_flip() +
  ggtitle("Boxplot of ERA for Starters vs Relievers") +
  xlab("Pitcher Type (Starter vs Reliever)")
box_plot_era
ggsave("boxplot_era.png", plot = last_plot(), width = 10, height = 6, dpi = 300)
```

# PCA

### Apply PCA and visualize it

```{r}
pca <- prcomp(pitching_df.num, scale = T)

data.frame(z1=-pca$x[,1],z2=pca$x[,2]) %>% 
  ggplot(aes(z1,z2,label=pitching_df$playerFullName, color=as.numeric(pitching_df$battersFaced)))+
  geom_point(size=0) +
  labs(title="PCA", x="PC1", y="PC2", color="Batters Faced") +
  theme_bw() +
  ylim(-5,5) +
  geom_text(size=2, hjust=0.6, vjust=0, check_overlap = TRUE)+
  scale_color_gradient(low = "blue", high = "red")

# Dimension 1 is starting pitchers
ggplot(data.frame(variable = names(pca$rotation[, 1]), loading = pca$rotation[, 1]), 
       aes(x = reorder(variable, loading), y = loading)) +
  geom_bar(stat = "identity", fill = "darkblue") +
  coord_flip() +
  theme_minimal() +
  labs(x = "Variable", y = "Loading on PC1", title = "Loadings for Principal Component 1") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) # Optional: rotate axis labels

# Dimension 2 is relief pitchers/closers
ggplot(data.frame(variable = names(pca$rotation[, 2]), loading = pca$rotation[, 2]), 
       aes(x = reorder(variable, loading), y = loading)) +
  geom_bar(stat = "identity", fill = "darkblue") +
  coord_flip() + 
  theme_minimal() +
  labs(x = "Variable", y = "Loading on PC2", title = "Loadings for Principal Component 2") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) # Optional: rotate axis labels

fviz_contrib(pca, choice = "var", axes = 1)
fviz_contrib(pca, choice = "var", axes = 2)

fviz_pca_var(pca, col.var = "contrib")
fviz_pca_biplot(pca, repel = TRUE, col.var = "contrib", geom = "point",  pointshape = 21, pointsize = 2, label = "all")
```

### Plot of the proportion of variance explained by each principal component including the cumulative proportion

```{r}
# Calculate proportion of variance explained
var_explained = pca$sdev^2 / sum(pca$sdev^2)
# Create a data frame with the results
summary_df = data.frame(
  PC = paste0("PC", 1:length(var_explained)),
  Proportion = var_explained,
  Cumulative = cumsum(var_explained)
)
# Plot the proportion of variance explained and the cummulative proportion
ggplot(head(summary_df, 9), aes(x = PC, y = Proportion)) +
  geom_col(fill = "steelblue") +
  geom_line(aes(y = Proportion), group = 1, color = "black") +
  geom_point(aes(y = Proportion), color = "black") +
  geom_line(aes(y = Cumulative), group = 1, color = "red") +
  geom_text(aes(y = Cumulative, label = scales::percent(Cumulative, accuracy = 0.1)), 
            vjust = -0.5, size = 3) +  # Adds text for line peaks
  geom_point(aes(y = Cumulative), color = "red") +
  geom_text(aes(label = scales::percent(Proportion, accuracy = 0.1)), 
            vjust = -0.5, size = 3) +  # Adds text for bar peaks
  
  scale_y_continuous(labels = scales::percent,
                     sec.axis = sec_axis(~., name = "Cumulative Proportion")) +
  labs(title = "Variance Explained by Principal Components",
       x = "Principal Component",
       y = "Proportion of Variance Explained") +
  theme_minimal()
```

### Find optimal loadings

Print the top 10 players

```{r}
pitching_df$playerFullName[order(pca$x[,1])][(nrow(pitching_df)-5):nrow(pitching_df)]
pitching_df$playerFullName[order(pca$x[,1])][1:10]
```

### Show the number of components to keep using Kaiser's Criterion

We can use Kaiser's Criterion that says the loadings that could be kept have a eigen value \> 1 such that the component explains more variance than a single original variable in the data set

```{r}
eigenvalues <- pca$sdev^2
components_to_keep <- sum(eigenvalues > 1)
components_to_keep 
```

8 eigen values can be kept as they have an eigen value \> 1

# Factor Analysis

Direct Factor Analysis did not work so we had to increase the tolerance for factor analysis, this is likely because this is a very noisy data set. We did it by removing the highly correlated variables

```{r}
ev <- eigen(cor(pitching_df.num))
nS <- nScree(x=ev$values)
plotnScree(nS,legend = T)
print(ev$values)

nzv <- nearZeroVar(pitching_df.num, saveMetrics = TRUE)
nzv
pitching_df_reduced <- pitching_df.num[, !nzv$nzv]

# Scale the pitching_df
pitching_df_scaled <- scale(pitching_df_reduced)

# Remove highly correlated variables with a cutoff of 0.95
cor_matrix <- cor(pitching_df_scaled)
high_cor <- findCorrelation(cor_matrix, cutoff = 0.95)
pitching_df_reduced <- pitching_df_scaled[, -high_cor]

fit <- factanal(pitching_df_reduced, factors = 3, rotation = "varimax", n.obs = nrow(pitching_df.num) * 1.5)
fit
```

## Factor Analysis with PCA

We used the first 10 components from the PCA to perform factor analysis

```{r}
fa_pca <- prcomp(pitching_df.num, scale = T)
summary(fa_pca)

# Pick the first 10 components
compnets_for_fa <- fa_pca$x[, 1:10]
fa <- factanal(compnets_for_fa, factors = 5, rotation = "varimax")
fa

# Scree plot to determine number of factors using Kaiser rule
eigenvalues <- fa_pca$sdev^2

plot(eigenvalues, type = "b", 
     # Smaller circle
     pch = 0,
     xlab = "Principal Component", 
     ylab = "Eigenvalue", 
     main = "Scree Plot")
abline(h = 1, col = "red", lty = 2) # Kaiser rule threshold line at eigenvalue = 1

```

### Graph the loadings

```{r}
loadings_df <- data.frame(Variable = rownames(fa$loadings))

for(i in 1:ncol(fa$loadings)) {
    loadings_df[paste0("Factor ", i)] <- fa$loadings[,i]
}

loadings_long <- melt(loadings_df, 
                     id.vars = "Variable",
                     variable.name = "Factor",
                     value.name = "Loading")

ggplot(loadings_long, aes(x = Variable, y = Loading)) +
  geom_col(fill = "lightgreen", width = 0.7) +
  facet_grid(Factor~.) +
  geom_text(aes(label = round(Loading, 2)), 
            color = "black",
            vjust = 1.3,
            size = 2.5) +
  labs(title = "Factor Loadings",
       x = "Component",
       y = "Loading") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.y = element_blank()
)
```

# Clustering

## KMEANS

```{r}
# First how much centers should we have 
fviz_nbclust(pitching_df.num, kmeans, method = 'silhouette', k.max = 10) # -> shows 2 clusters is optimal
k = 3 # -> I chose 3 anyway
fit =  kmeans(scale(pitching_df.num), k, nstart=100)
groups = fit$cluster
# How many players are in each cluster
barplot(table(groups), col="darkblue")
# Create unique labels by appending an index to each duplicate name
unique_labels <- make.unique(as.character(pitching_df$playerFullName))
rownames(pitching_df.num) <- unique_labels
fviz_cluster(fit, pitching_df.num, geom = "text", labelsize=8, main = "Clusters with K-means")
# More readable plot
fviz_cluster(fit, data = pitching_df.num, geom = "point", label = "none", main = "Clusters with K-means") + geom_text(label = unique_labels, check_overlap = T, size = 3)

```

### Interpretation of centers:

Plot of the contribution of each variable to each cluster

```{r, fig.align='center', fig.width=10}
centers=fit$centers
tidy = cbind(
  gather(as.data.frame(t(centers)), "cluster", "coor"),
  var=rep(colnames(centers, k)),
  size=rep(table(fit$cluster), each=ncol(centers))
  )

tidy %>%
  ggplot(aes(x=cluster, y=coor, fill=cluster)) +
  geom_col() +
  facet_wrap(~var) +
  geom_text(aes(label=size),position=position_stack(1.2))
```

Different plot also showing how each variable contributes to the clusters

```{r, fig.align='center', fig.width=10}
centers=fit$centers
tidy = cbind(
  gather(as.data.frame(t(centers)), "cluster", "coor"),
  var=rep(colnames(centers, k)),
  size=rep(table(fit$cluster), each=ncol(centers)) 
  )

tidy %>%
  ggplot(aes(x=var, y=coor, fill=var)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~cluster) +
  theme(axis.text.x = element_text(angle=45),legend.position = "none")
```

Showing clusters individually

```{r}
# Use the colors as in fviz_cluster
colors <- c("#fc7671", "#00bb46", "#5d9bfa")
groups = fit$cluster
lapply(1:3, function(cluster_id) {
  # Subset data
  subset_data <- pitching_df.num[groups == cluster_id, ]

  # Re-run clustering and plot
  fit_subset <- kmeans(scale(subset_data), centers = 1, nstart = 50)
    p2 <- fviz_cluster(fit_subset, 
                  data = subset_data, 
                  geom = c("point", "text"),
                  labelsize = 6,
                  repel = F,
                  ellipse.type = "convex",
                  ggtheme = theme_minimal()) +
                  scale_color_manual(values = colors[cluster_id]) +
                  theme(legend.position = "none") +
                  ggtitle(paste("Cluster", cluster_id))
})
```



A silhouette plot to show how well the clusters are separated Eclust scalss the data automatically for us

```{r}
fit.kmeans = eclust(pitching_df.num, "kmeans", k=3, stand=TRUE, nstart=100, graph = FALSE)
fviz_silhouette(fit.kmeans)
```

### Choosing the number of clusters for kmeans

WSS method Silhouette method Gap statistic method

```{r}
# WSS
fviz_nbclust(scale(pitching_df.num), kmeans, method = "wss", k.max = 20, nstart = 200, iter.max = 100, nboot = 100)
# Silhouette
fviz_nbclust(scale(pitching_df.num), kmeans, method = "silhouette", k.max = 20, nstart = 1000, iter.max = 100)
# Gap statistic
fviz_nbclust(scale(pitching_df.num), kmeans, method = 'gap_stat', k.max = 7, nstart = 20, nboot = 100, iter.max = 500)

```

# Kmeans of PCA

Does the \# of clusters correlate to without pca?

No the optimal still says 2

# WSS method

```{r}

cluster_pca <- prcomp(scale(pitching_df.num), scale = T)

my_pca_pitching_df <- data.frame(cluster_pca$x[, 1:2])

fviz_nbclust(my_pca_pitching_df, kmeans, method = 'silhouette', k.max = 10)

my_kmeans <- kmeans(my_pca_pitching_df, centers = 3)

rownames(my_pca_pitching_df) <- unique_labels
fviz_cluster(my_kmeans, data = my_pca_pitching_df, geom = "text", labelsize=8)

centers=my_kmeans$centers
tidy = cbind(
  gather(as.data.frame(t(centers)), "cluster", "coor"),
  var=rep(colnames(centers, k)),
  size=rep(table(my_kmeans$cluster), each=ncol(centers))
)

tidy %>%
  ggplot(aes(x=cluster, y=coor, fill=cluster)) +
  geom_col() +
  facet_wrap(~var) +
  geom_text(aes(label=size),position=position_stack(1.2))

centers=my_kmeans$centers
tidy = cbind(
  gather(as.data.frame(t(centers)), "cluster", "coor"),
  var=rep(colnames(centers, k)),
  size=rep(table(my_kmeans$cluster), each=ncol(centers))
  )

tidy %>%
  ggplot(aes(x=var, y=coor, fill=var)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~cluster) +
  theme(axis.text.x = element_text(angle=45),legend.position = "none")

```

## PAM clustering

```{r}
fit.pam = eclust(pitching_df.num, FUNcluster="pam", stand = TRUE, k=3,
                 graph = T, nstart=1000)

fviz_cluster(fit.pam, geom="point", main="Clusters with PAM", ellipse.type = "norm")+
geom_text(label=pitching_df$playerFullName, check_overlap = T)
```

### Number of clusters for PAM

```{r n-groups-pam-sil}
# WSS
fviz_nbclust(scale(pitching_df.num), pam, method="wss", k.max=10)
# Silhouette
fviz_nbclust(scale(pitching_df.num), pam, method="silhouette", k.max=10)
# Gap statistic
fviz_nbclust(scale(pitching_df.num), pam, method="gap_stat", k.max=10)
```

## Adjusted rand index

```{r}
adjustedRandIndex(fit.kmeans$cluster, fit.pam$clustering)
```

## Kernel Estimation

```{r}
create_kde_plot <- function(pitching_df, col_name) {
  p <- ggplot(pitching_df, aes_string(x = col_name)) +
    geom_density(fill = "darkblue", color = "black", alpha = 0.7) +
    labs(title = paste("Kernel Density Estimation of", col_name), 
         x = col_name, y = "Density") +
    theme_minimal()
  
  print(p)
}

for (col_name in names(pitching_df.num)) {
  if (is.numeric(pitching_df.num[[col_name]])) {
    create_kde_plot(pitching_df.num, col_name)
  }
}

```
